---
title: "AlberdiLab | Aizpurua et al. 2024"
subtitle: "Calotriton metagenomics"
author:
  - Ostaizka Aizpurua^[University of Copenhagen, ostaizka.aizpurua@sund.ku.dk]
  - Carlos Cabido^[Aranzadi Science Society]
  - Garazi Martin Bideguren^[University of Copenhagen, garazi.bideguren@sund.ku.dk]
  - Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/calotriton_metagenomics
description: |
  Data analysis code for the study on the study of gut microbiomes of Pyrenean brook salamanders in high and low altitudes.
link-citations: yes
github-repo: alberdilab/calotriton_metagenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across different rivers.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/calotriton_metagenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(ANCOMBC)
library(lme4)
```

<!--chapter:end:index.Rmd-->

# Prepare data

## Load data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
sample_metadata <- read_tsv("data/metadata_antibiotics.tsv")
```

### Read counts

```{r load_read_counts_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts_raw <- read_tsv("data/genome.count.tsv") %>%
    rename(genome=1)

#Transformation of read_counts to combine data from both sequence rounds
merge_and_rename <- function(read_counts_raw) {
  read_counts_raw %>%
    # Gather the columns into long format
    pivot_longer(cols = -genome, names_to = "col") %>%
    # Extract prefix
    mutate(prefix = gsub("^(.*?)_.*", "\\1", col)) %>%
    # Group by prefix and genome, then summarize
    group_by(prefix, genome) %>%
    summarize(value = sum(value)) %>%
    # Spread the data back to wide format
    pivot_wider(names_from = prefix, values_from = value)
}

read_counts <- merge_and_rename(read_counts_raw)
```

### Genome base hits

```{r load_genome_hits_1, warning=FALSE, comments="", message=FALSE}
genome_hits_raw <- read_tsv("data/genome.covered_bases.tsv") %>%
    rename(genome=1)

#Transformation of genome_hits to combine data from both sequence rounds
merge_and_rename <- function(genome_hits_raw) {
  genome_hits_raw %>%
    # Gather the columns into long format
    pivot_longer(cols = -genome, names_to = "col") %>%
    # Extract prefix
    mutate(prefix = gsub("^(.*?)_.*", "\\1", col)) %>%
    # Group by prefix and genome, then summarize
    group_by(prefix, genome) %>%
    summarize(value = sum(value)) %>%
    # Spread the data back to wide format
    pivot_wider(names_from = prefix, values_from = value)
}

genome_hits <- merge_and_rename(genome_hits_raw)
```

### Genome taxonomy

```{r load_genome_metadata_1, warning=FALSE, comments="", message=FALSE}
genome_taxonomy <- read_tsv("data/gtdbtk.summary.tsv") %>%
  select(mag_id = user_genome, classification) %>%
  separate(
    classification,
    into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";") %>%
    rename(genome=1)
```

### Genome quality

```{r load_genome_quality, warning=FALSE, comments="", message=FALSE}
genome_quality <- read_tsv("data/quality_report.tsv") %>%
  select(
    genome = Name, 
    completeness = Completeness, 
    contamination = Contamination,
    length = Genome_Size, 
    gc = GC_Content
  )
genome_quality<-genome_quality %>% 
  mutate (genome=str_remove_all(genome,".fa"))
```

### Genome tree

```{r load_genome_tree_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("data/gtdbtk.backbone.bac120.classify.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_taxonomy$genome) # keep only MAG tips
```

### Genome annotations

```{r load_genome_annotations_1, warning=FALSE, comments="", message=FALSE}
genome_annotations <- read_tsv("data/annotations.tsv.gz") %>%
    rename(gene=1, genome=2, contig=3)
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Merge genome taxonomy and quality

```{r generate_genome_metadata, warning=FALSE, comments="", message=FALSE}
genome_metadata <- genome_taxonomy %>%
  inner_join(genome_quality,by=join_by(genome==genome)) #join quality
```

### Calculate genome coverage

```{r calculate_genome_coverage, warning=FALSE, comments="", message=FALSE}
genome_coverage <- genome_hits %>%
  mutate(across(where(is.numeric), ~ ./genome_metadata$length))
```

### Filter reads by coverage

```{r filter_coverage_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_counts %>%
  semi_join(genome_quality, by = "genome")

min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]])) 
```

### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

### Distill annotations into GIFTs 

```{r distill_annotations_1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=2,annotcol=c(9,10,19))
```

## Load data statistics

### Raw reads

```{r raw_reads, warning=FALSE, comments="", message=FALSE}
raw_reads <-
  "data/multiqc_general_stats.txt" %>%
  read_tsv() %>%
  select(
    sample = Sample,
    raw_reads = `FastQC_mqc-generalstats-fastqc-total_sequences`
  ) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(raw_reads = sum(raw_reads, na.rm = TRUE)) %>%
  rename(sample = sample_prefix)  %>%
  mutate(sample = str_remove(sample, "^fastp \\|\\s*"))
```

### Quality-filtered reads

```{r filtered_reads, warning=FALSE, comments="", message=FALSE}
fastp_reads <-
  "data/multiqc_general_stats.txt" %>%
  read_tsv() %>%
  filter(str_detect(Sample, "fastp")) %>%
  select(
    sample = Sample,
    trimmed_reads = `FastQC_mqc-generalstats-fastqc-total_sequences`
  ) %>%
  mutate(
    sample =
      sample %>%
        str_remove_all("_[u12]+$") %>%
        str_remove_all("^fastp \\| ")
  ) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(trimmed_reads = sum(trimmed_reads, na.rm = TRUE)) %>%
  rename(sample = sample_prefix)
```

### Host-mapped reads

```{r host_mapped, warning=FALSE, comments="", message=FALSE}
host_mapped <-
  "data/multiqc_general_stats.txt" %>%
  read_tsv() %>%
  filter(!str_detect(Sample, "fastp")) %>%
  select(
    sample = Sample,
    host_mapped = `Samtools_mqc-generalstats-samtools-reads_mapped`,
    mapping_total = `Samtools_mqc-generalstats-samtools-raw_total_sequences`
  ) %>%
  mutate(
    host_unmapped = mapping_total - host_mapped
  ) %>%
  filter(!is.na(host_mapped)) %>%
  separate(
    col = sample,
    into = c("host_name", "sample"),
    sep = " \\| "
  ) %>%
  rename(mapped = host_mapped, unmapped = host_unmapped) %>%
  select(-mapping_total) %>%
  pivot_longer(-host_name:-sample) %>%
  mutate(
    sample =
      sample %>%
        str_remove_all("_[u12]+$") %>%
        str_remove_all("^fastp \\| ")
  ) %>%
  select(-host_name) %>%
  pivot_wider()%>%
  mutate(sample = str_extract(sample, "^[^_]+")) %>%
  group_by(sample) %>%
  summarize(
    mapped = sum(mapped),
    unmappedd = sum(unmapped)
  )
```

### Prokaryotic fraction

```{r singlem, warning=FALSE, comments="", message=FALSE}
singlem <-
  "data/microbial_fraction.tsv" %>%
  read_tsv() %>%
  distinct() %>%
  mutate(
    sample = sample %>% str_remove_all("_1$"),
   read_fraction = read_fraction %>% str_remove("%") %>% as.numeric(),
   read_fraction = read_fraction / 100
  ) %>%
  select(
   sample,
   singlem_prokaryotic_bases = bacterial_archaeal_bases,
   singlem_metagenome_size = metagenome_size,
   singlem_read_fraction = read_fraction,
  ) %>%
mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
  group_by(sample_prefix) %>%
  summarize(
    singlem_prokaryotic_bases = sum(singlem_prokaryotic_bases),
    singlem_metagenome_size = sum(singlem_metagenome_size),
    singlem_read_fraction = mean(singlem_read_fraction)
  ) %>%
  rename(sample = sample_prefix)
```

### MAG-mapped reads

```{r mag_mapping, warning=FALSE, comments="", message=FALSE}
mag_mapping <-
  read_tsv("data/contig.count.tsv") %>%
  pivot_longer(-sequence_id) %>%
  summarise(value = sum(value), .by = "name") %>%
  rename(sample = name, mapped_mags = value) %>%
  mutate(sample_prefix = str_extract(sample, "^[^_]+")) %>%
    group_by(sample_prefix) %>%
    summarize(
      mapped_mags = sum(mapped_mags)
    ) %>%
    rename(sample = sample_prefix)
```

### Wrap data statistics

```{r wrap_statistics, warning=FALSE, comments="", message=FALSE}
data_stats <- raw_reads %>%
  left_join(fastp_reads) %>%
  left_join(host_mapped) %>%
  left_join(singlem) %>%
  left_join(mag_mapping)

data_stats<- data_stats %>%
  filter(!str_detect(sample, "newr \\|")) %>%
  filter(!str_detect(sample, "nonnewr \\|"))
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     genome_gifts, 
     phylum_colors,
     data_stats,
     file = "data/data.Rdata")
```

<!--chapter:end:01_data_preparation.Rmd-->

# Data statistics

```{r load_data_stats}
load("data/data.Rdata")
```

## Sequencing reads statistics

```{r reads_stats}
data_stats$raw_reads %>% sum()
data_stats$raw_reads %>% mean()
data_stats$raw_reads %>% sd()
```

## DNA fractions
```{r data_fractions_stats}
#Overall
data_stats %>%
    mutate(mapped_perc=mapped_mags/trimmed_reads) %>%
    summarise(mean=mean(mapped_perc),sd=sd(mapped_perc))
```


```{r data_fractions_plot, message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
data_stats %>%
  mutate(
    low_quality = raw_reads - trimmed_reads,
    unmapped_reads = trimmed_reads - mapped - mapped_mags
  ) %>%
  select(sample, low_quality, mapped, mapped_mags, unmapped_reads) %>%
  pivot_longer(-sample) %>%
  mutate(name=factor(name,levels=c("low_quality","mapped","unmapped_reads","mapped_mags"))) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(!is.na(time_point)) %>%
  ggplot(aes(x = sample, y = value, fill = name)) +
      geom_bar(stat = "identity", position = "fill") +
      scale_fill_manual(name="Sequence type",
                    breaks=c("low_quality","mapped","unmapped_reads","mapped_mags"),
                    labels=c("Low quality","Mapped to host","Unmapped","Mapped to MAGs"),
                    values=c("#CCCCCC", "#bcdee1", "#d8b8a3","#93655c"))+
      facet_grid(~time_point, scales = "free") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, size=0)) +
      labs(y="DNA sequence fraction",x="Samples")
```

## Recovered microbial fraction

```{r data_estimations_plot, message=FALSE, warning=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
data_stats %>%
  mutate(
    unmapped_reads = trimmed_reads - mapped - mapped_mags,
    mag_proportion = mapped_mags / (mapped_mags + unmapped_reads),
    singlem_read_fraction = singlem_read_fraction
  ) %>%
  select(sample, mag_proportion, singlem_read_fraction) %>%
  mutate(
    mag_proportion = if_else(singlem_read_fraction == 0, 0, mag_proportion),
    singlem_read_fraction = if_else(singlem_read_fraction == 0, NA, singlem_read_fraction),
    singlem_read_fraction = if_else(singlem_read_fraction < mag_proportion, NA, singlem_read_fraction),
    singlem_read_fraction = if_else(singlem_read_fraction > 1, 1, singlem_read_fraction)
  ) %>%
  pivot_longer(-sample, names_to = "proportion", values_to = "value") %>%
  mutate(
    proportion = factor(
      proportion,
      levels = c("mag_proportion", "singlem_read_fraction")
    )
  ) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(!is.na(time_point)) %>%
  ggplot(aes(x = sample, y = value, color = proportion)) +
      geom_line(aes(group = sample), color = "#f8a538") +
      geom_point() +
      scale_color_manual(name="Proportion",
                    breaks=c("mag_proportion","singlem_read_fraction"),
                    labels=c("Recovered","Estimated"),
                    values=c("#52e1e8", "#876b53"))+
      theme_minimal() +
      facet_grid(~time_point, scales = "free", space="free")+
      labs(y = "Samples", x = "Prokaryotic fraction") +
      scale_y_continuous(limits = c(0, 1)) +
      theme(
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text( angle = 90, vjust = 0.5, hjust = 1, size = 0),
        legend.position = "right"
      )
```

```{r damr, message=FALSE, warning=FALSE}
damr <- data_stats %>%
  mutate(
    unmapped_reads = trimmed_reads - mapped - mapped_mags,
    mag_proportion = mapped_mags / (mapped_mags + unmapped_reads),
    singlem_read_fraction = singlem_read_fraction
  ) %>%
  mutate(damr=pmin(1, mag_proportion/singlem_read_fraction)) %>%
  select(sample,damr)

damr %>% 
  tt()

damr %>% 
  filter(damr>95) %>%
  nrow()

damr %>% 
  summarise(mean=mean(damr),sd=sd(damr)) %>% 
  tt()
```

<!--chapter:end:02_data_statistics.Rmd-->

# MAG catalogue

```{r load_data_mag}
load("data/data.Rdata")
```

## Genome phylogeny

```{r genome_phylogeny, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.5)

# Add phylum ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.55, width=0.1, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors) +
        geom_tiplab2(size=1, hjust=-0.1) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=genome_metadata,
                geom=geom_bar,
                mapping = aes(x=completeness, y=genome, fill=contamination),
                offset = 0.55,
                orientation="y",
              stat="identity")

# Add genome-size ring
circular_tree <-  circular_tree +
        new_scale_fill() +
        scale_fill_manual(values = "#cccccc") +
        geom_fruit(
             data=genome_metadata,
             geom=geom_bar,
             mapping = aes(x=length, y=genome),
                 offset = 0.05,
                 orientation="y",
         stat="identity")

# Add text
circular_tree <-  circular_tree +
        annotate('text', x=2.7, y=0, label='             Phylum', family='ArialMT', size=3.5) +
        annotate('text', x=3.1, y=0, label='                     Genome quality', family='ArialMT', size=3.5) +
        annotate('text', x=3.5, y=0, label='                  Genome size', family='ArialMT', size=3.5)

#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```
## Taxonomy overview

```{r genome_taxonomy_bacteria}
genome_metadata %>% 
  group_by(phylum) %>%
  summarise(mag_n=n()) %>%
  arrange(-mag_n) %>%
  tt()
```

## Genome quality

```{r genome_quality}
genome_metadata %>% 
    summarise(completeness_mean=mean(completeness) %>% round(2) %>% as.character(), 
              completeness_sd=sd(completeness) %>% round(2) %>% as.character(), 
              contamination_mean=mean(contamination) %>% round(2), 
              contamination_sd=sd(contamination) %>% round(2)) %>%
    unite("Completeness",completeness_mean, completeness_sd, sep = " ± ", remove = TRUE) %>%
    unite("Contamination",contamination_mean, contamination_sd, sep = " ± ", remove = TRUE) %>%
    tt()
```

```{r genome_quality_plot, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

#Generate quality biplot
genome_biplot <- genome_metadata %>%
  select(c(genome,domain,phylum,completeness,contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=completeness,y=contamination,size=length,color=phylum)) +
              geom_point(alpha=0.7) +
                    ylim(c(10,0)) +
                    scale_color_manual(values=phylum_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate contamination boxplot
genome_contamination <- genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(50,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))

```


## Functional overview

```{r function_heatmap, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db)

# Generate  basal tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(., size = 0.3) 

#Add phylum colors next to the tree tips
function_tree <- gheatmap(function_tree, phylum_heatmap, offset=0, width=0.1, colnames=FALSE) +
            scale_fill_manual(values=phylum_colors) +
            labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

#Add functions heatmap
function_tree <- gheatmap(function_tree, function_table, offset=0.5, width=3.5, colnames=FALSE) +
            vexpand(.08) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white") +
            labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

# Add completeness barplots
function_tree <- function_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\ncompleteness")

function_tree
```

## Functional ordination

```{r function_ordination, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
# Generate the tSNE ordination
tSNE_function <- Rtsne(X=function_table, dims = 2, check_duplicates = FALSE)

# Plot the ordination
function_ordination <- tSNE_function$Y %>%
                as.data.frame() %>%
                mutate(genome=rownames(function_table)) %>%
                inner_join(genome_metadata, by="genome") %>%
                rename(tSNE1="V1", tSNE2="V2") %>%
                select(genome,phylum,tSNE1,tSNE2, length) %>%
                ggplot(aes(x = tSNE1, y = tSNE2, color = phylum, size=length))+
                            geom_point(shape=16, alpha=0.7) +
                            scale_color_manual(values=phylum_colors) +
                            theme_minimal() +
                labs(color="Phylum", size="Genome size") +
                guides(color = guide_legend(override.aes = list(size = 5))) # enlarge Phylum dots in legend

function_ordination
```

<!--chapter:end:03_mag_catalogue.Rmd-->

# Community composition

```{r load_data_community}
load("data/data.Rdata")
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~ individual + time_point,  scales="free", labeller=labeller(time_point=c("Pre_antibiotics"="Pre_ant", "Post_antibiotics_1"="Post_ant_1", "Post_antibiotics_2"="Post_ant_2")) ) + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum) %>%
  summarise(relabun=sum(count))

phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) %>%
    tt()
```

```{r taxonomy_boxplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    filter(phylum %in% phylum_arrange) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        theme(legend.position="none") +
        labs(y="Phylum",x="Relative abundance")
```

## Taxonomy boxplot

### Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    group_by(family) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) %>%
    tt()
```


```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    pull()

# Per time_point
family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~time_point)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```

### Genus

```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__") %>%
  mutate(genus= sub("^g__", "", genus))

genus_summary_sort <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun, na.rm=T),sd=sd(relabun, na.rm=T)) %>%
    arrange(-mean) 

genus_summary_sort %>%
    tt()
```

```{r taxonomy_jitterplot_genus, fig.height=14, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

#Per time_point
genus_summary %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus=factor(genus, levels=rev(genus_summary_sort %>% pull(genus)))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~time_point)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```


<!--chapter:end:04_community_composition.Rmd-->

# Alpha diversity

```{r load_data_alpha}
load("data/data.Rdata")
```

```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% rownames(dist)) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

```{r alpha_div_boxplot, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=10, fig.fullwidth=TRUE}
#Richness
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(metric=="richness") %>%
      ggplot(aes(y = value, x = time_point, group=time_point, color=time_point, fill=time_point)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )

#Neutral
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(metric=="neutral") %>%
      ggplot(aes(y = value, x = time_point, group=time_point, color=time_point, fill=time_point)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )

#Phylogenetic
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(metric=="phylogenetic") %>%
      ggplot(aes(y = value, x = time_point, group=time_point, color=time_point, fill=time_point)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )

#Functional
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(metric=="functional") %>%
      ggplot(aes(y = value, x = time_point, group=time_point, color=time_point, fill=time_point)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1) +
  scale_x_discrete(name ="Time-point", 
                    limits=c("2","1","0.5"))
      )
```



<!--chapter:end:05_alpha_diversity.Rmd-->

# Beta diversity

```{r load_data_beta}
load("data/data.Rdata")
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

### Permanova

```{r permanova, comment="", message=FALSE, warning=FALSE, eval=FALSE}
#Richness
betadisper(beta_q0n$S, sample_metadata$time_point) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q0n$S ~ time_point, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Neutral diversity
betadisper(beta_q1n$S, sample_metadata$time_point) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1n$S ~ time_point, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Phylogenetic diversity
betadisper(beta_q1p$S, sample_metadata$time_point) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1p$S ~ time_point, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Functional diversity
betadisper(beta_q1f$S, sample_metadata$time_point) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1f$S ~ time_point, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```

### Richness diversity plot

```{r beta_div_nmds_richness_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(time_point) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = time_point, fill = time_point, shape = as.factor(individual))) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```


### Neutral diversity plot

```{r beta_div_nmds_neutral_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(time_point) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = time_point, fill = time_point, shape = as.factor(individual))) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

### Phylogenetic diversity plot

```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q1p$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(time_point) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = time_point, fill = time_point, shape = as.factor(individual))) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

### Functional diversity plot

```{r beta_div_nmds_func_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q1f$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(time_point) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = time_point, fill = time_point, shape = as.factor(individual))) +
      scale_color_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC9900","#66CC99", "#990066")) +
      scale_fill_manual(name="Time_point",
          breaks=c("Pre_antibiotics","Post_antibiotics_1", "Post_antibiotics_2"),
          labels=c("Pre_ant","Post_ant_1", "Post_ant_2"),
          values=c("#CC990050","#66CC9950", "#99006650")) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

<!--chapter:end:06_beta_diversity.Rmd-->

